VFTS682: a confirmed dynamical ejection?

### 2018-05-10 VFTS682.tex
re-read stuff

\author{M.~Renzo\inst{1} \and \todo{TBD}% S.~E.~de~Mink\inst{1} \and
  % S.~Justham\inst{2,3} \and A.~de~Koter\inst{1} \todo{etc..order to be decided}
  .}

@@

The second data release (DR2) from the Gaia satellite
\cite[][]{gaia:16,brown:18} allows us to test
these hypothesis using one particular star, VFTS682. 
This star belongs to the 30 Doradus region in the Large Magellanic
Cloud (LMC), and has an inferred initial mass of $M_\mathrm{ZAMS}=150.0^{+28.7}_{-17.4}\,M_\odot$
\citep[][]{bestenlehner:11,schneider:18}. Its spectral type is WNh5, and it is presently observed at a
projected distance of $\sim$$29$\,pc from the nearest cluster of
massive stars R136 \citep[also known as NGC2070][]{bestenlehner:11}. The inferred mass loss
rate is $\sim10^{-XX}\,M_\odot \ \mathrm{yr}^{-1}$ \todo{ref}.

@@

We retrieve for VFTS682 the position in right ascension (RA) and declination (DE)
in the IGCS frame \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DE}$,
respectively). For the radial velocity of VFTS682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

@@

  \begin{tabular}[htbp]{l|c|c}
    Parameter & Value & Source\\ \hline\hline
    RA \hfill[degree] &  \phantom{-}84.73 $\pm$  0.03 & \multirow{4}{*}{Gaia DR2}\\
    DE \hfill [degree] & -69.07 $\pm$  0.05  & \\
    $\mu_\mathrm{RA}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}1.84 $\pm$ 0.07 & \\
    $\mu_\mathrm{DE}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}0.78 $\pm$ 0.08& \\
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & \phantom{0}-45 $\pm$ 25 & \cite{bestenlehner:11}\\
    \hline
  \end{tabular}
  \label{tab:vfts682}
\end{table}

@@

Using the 651 stars selected as described in \Secref{sec:sample} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}^\mathrm{sur}\rangle = 1.683\pm0.002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DE}^\mathrm{sur}\rangle =
0.672\pm0.003\,\mathrm{mas\ yr^{-1}}$. We note that these values are
in good agreement with what found by \cite{lennon:18}. Subtracting these values from the
proper motions of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu^\mathrm{sur}_\mathrm{RA} = 0.16\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu^\mathrm{sur}_\mathrm{DE} =
0.11\pm 0.08\,\mathrm{mas\ yr^{-1}}$. We note that the error budget is
dominated by the errors on the proper motion components of VFTS682.

These can be converted in the
the components of the transverse velocity $v^\mathrm{sur}_\mathrm{RA}=38\pm17\,\kms$,
$v^\mathrm{sur}_\mathrm{DE}=26\pm19\,\kms$, assuming a distance of
50\,kpc (we do not account for the uncertainty in the distance
estimate when propagating errors). The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, allowing us to calculate the speed of the star:

\begin{equation}
  \label{eq:speed_around}
  v^\mathrm{sur} = \sqrt{\left(v^\mathrm{sur}_\mathrm{RA}\right)^2
    +\left(v^\mathrm{sur}_\mathrm{DE}\right)^2+\left(\delta
      v_\mathrm{rad}\right)^2} = 64 \pm 21 
  \, \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS682 with respect the
surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} =
  \frac{29\,\mathrm{pc}}{46\,\kms} \simeq 0.63\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel = 46\pm
17\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}, and we use the approximation $1 \kms \simeq 1
\mathrm{pc \ Myr^{-1}}$.

@@

Based on our results, we can claim that VFTS682 is the most massive
runaway known to date, with a peculiar three-dimensional spatial
velocity of $\sim60\,\kms$. This means that isolated star formation is
\emph{not} required to explain this star. Its proper motion suggests that it was ejected from the cluster R136
$\sim0.6$\,Myr ago. Given the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

Dynamical ejections due to N-body interactions typically (although not necessarily) eject the least
massive star among those interacting \todo{ref}. This means that, just
based on the kinematic properties of VFTS682, we would expect several
stars with initial masses larger than $\sim150\,M_\odot$ in R136.
This can be considered an independent confirmation of the detection
of extremely massive stars by \cite{crowther:10} in the core of the
cluster. The projected rotational equatorial velocity of VFTS682
reported by \cite{schneider:18} is $v\sin(i)<200\,\kms$, which is in
line with the average rotation rate of massive stars in the region
\citep[][]{ramirez-agudelo:15}. This suggests that VFTS682 has not
experienced binary interactions, nor it will, since the multi-epoch
data of the VFTS survey rule out present-day binarity. Moreover, the spectral type of VFTS682
\citep[WNh5][]{bestenlehner:11} is the same as R136a1-a3, i.e.~the
three most massive stars detected in the core of the cluster by
\cite{crowther:10}. Therefore,
VFTS682 might be an ideal target to constrain the stellar physics of
stars with masses well above $\sim100\,M_\odot$: its isolation makes
it an easier target for observation than the crowded core of R136,
where similar stars exist.

The similarities between VFTS682 and the WNh5 stars in the core of
R136 are also in agreement with the ``bully binary'' model of
\cite{fujii:11}. They suggested, based on their numerical simulations,
that dynamical interactions lead to the formation of an extremely
massive binary in the early phases of a cluster evolution, and then it tightens by ejecting other stars passing
by. Interpreting our results for VFTS682 through the lens of their simulations would
suggest the presence of a binary with total mass
$M_1+M_2\gtrsim 400\,M_\odot$ in the core of the cluster, which they
suggested could be R145. The kinematic age of VFTS682 puts an
upperlimit to the timescale to form such ``bully binary'' in
R136. Unfortunately, their simulations did not include stars more
massive than $100\,M_\odot$, so it is difficult to predict, based on
the existence of VFTS682, how many other stars should have been
ejected from the cluster already and what is their mass
distribution. A comprehensive study of the kinematic properties of the
large sample of stars with extremely large inferred masses visible
around R136 is encouraged.

\cite{lennon:18} have carried out a similar study on VFTS16, another
massive star ($\sim90\,M_\odot$) in the 30 Doradus region, previously
known to be a runaway from the value of
its radial velocity. They also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster. 
The value of $\tau_\mathrm{kin}\simeq0.63$\,Myr we find for VFTS682 (see \Secref{sec:r136}) is smaller
than the corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic age of
$\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

\begin{itemize}
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\item comment on final fate (BH)

\end{itemize}

@@

\begin{acknowledgements}
  The background image of \Figref{fig:main} is based on observations
  made with ESO Telescopes at the La Silla Observatory under programme
  ID 076.C-0888, processed and released by the ESO VOS/ADP group.
  This research made use of APLpy, an open-source plotting package for Python \cite{robitaille:12}.
  We are grateful to S.~Torres for extremely useful discussions on the
  Gaia DR2 dataset, and to M.~C.~Ramirez-Tannus for help with APLpy.
\end{acknowledgements}
\end{document}

>>>>

\author{M.~Renzo\inst{1} \and S.~E.~de~Mink\inst{1} \todo{etc...}%\and
  % S.~Justham\inst{2,3} \and A.~de~Koter\inst{1} \todo{etc..order to be decided}
  .}

@@

The second data release (DR2) from the Gaia satellite
\cite[][]{gaia:16,brown:18} allows us to test
these hypothesis using one particular star, VFTS682. 
This star belongs to the 30 Doradus region in the Large Magellanic
Cloud (LMC), and has an inferred initial mass of $M_\mathrm{ZAMS}=150.0^{+28.7}_{-17.4}\,M_\odot$
\citep[][]{bestenlehner:11,schneider:18}. Its spectral type is WNh5, and it is presently observed at a
projected distance of $\sim$$29$\,pc from the nearest cluster of
massive stars R136 \citep[also known as NGC2070][]{bestenlehner:11}. The inferred mass loss
rate is $\sim10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$ \todo{ref}.

@@

We retrieve for VFTS682 the position in right ascension (RA) and declination (DEC)
in the IGCS frame \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DEC}$,
respectively). For the radial velocity of VFTS682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

@@

  \begin{tabular}[htbp]{l|c|c}
    Parameter & Value & Source\\ \hline\hline
    RA \hfill[degree] &  \phantom{-}84.73 $\pm$  0.03 & \multirow{4}{*}{Gaia DR2}\\
    DEC \hfill [degree] & -69.07 $\pm$  0.05  & \\
    $\mu_\mathrm{RA}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}1.84 $\pm$ 0.07 & \\
    $\mu_\mathrm{DEC}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}0.78 $\pm$ 0.08& \\
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & \phantom{0}-45 $\pm$ 25 & \cite{bestenlehner:11}\\
    \hline
  \end{tabular}
  \label{tab:vfts682}
\end{table}

@@

Using the 651 stars selected as described in \Secref{sec:sample} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}\rangle = 1.683\pm0.002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DEC}\rangle =
0.672\pm0.003\,\mathrm{mas\ yr^{-1}}$. We note that these values are
in good agreement with what found by \cite{lennon:18}. Subtracting these values from the
proper motions of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu_\mathrm{RA} = 0.16\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu_\mathrm{DEC} =
0.11\pm 0.08\,\mathrm{mas\ yr^{-1}}$. We note that the error budget is
dominated by the errors on the proper motion components of VFTS682.

These can be converted in the
the components of the relative transverse velocity $\delta v_\mathrm{RA}=38\pm17\,\kms$,
$\delta v_\mathrm{DEC}=26\pm19\,\kms$, assuming a distance of
50\,kpc (we do not account for the uncertainty in the distance
estimate when propagating errors). The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, allowing us to calculate the three-dimensional
peculiar speed of the star:

\begin{equation}
  \label{eq:speed_around}
  v_\mathrm{pec} = \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2+\left(\delta
      v_\mathrm{rad}\right)^2} = 64 \pm 21 
  \, \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS682 with respect the
surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} =
  \frac{29\,\mathrm{pc}}{46\,\kms} \simeq 0.63\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =46\pm
17\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}, and we use the approximation $1 \kms \simeq 1
\mathrm{pc \ Myr^{-1}}$.

@@

Based on our results, we can claim that VFTS682 is the most massive
runaway known to date, with a peculiar three-dimensional spatial
velocity of $\sim$$60\,\kms$. This means that isolated star formation is
\emph{not} required to explain this star. Its proper motion suggests that it was ejected from the cluster R136
$\sim$$0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

Dynamical ejections due to N-body interactions typically (although, not necessarily) eject the least
massive star among those interacting \todo{ref}. This means that, just
based on the kinematic properties of VFTS682, we would expect several
stars with initial masses larger than $\sim$$150\,M_\odot$ in the
cluster R136.
This can be considered an independent confirmation of the detection
of extremely massive stars by \cite{crowther:10} in the core of the
cluster. The projected rotational equatorial velocity of VFTS682
reported by \cite{schneider:18} is $v\sin(i)<200\,\kms$, which is in
line with the average rotation rate of massive stars in the region
\citep[][]{ramirez-agudelo:15}. This suggests that VFTS682 has not
experienced binary interactions, nor it will, since the multi-epoch
data of the VFTS survey rule out the presence of a companion at
present day. Moreover, the spectral type of VFTS682
\citep[WNh5,][]{bestenlehner:11} is the same as R136a1-a3, i.e.~the
three most massive stars detected in the core of the cluster by
\cite{crowther:10}. Therefore,
VFTS682 might be an ideal target to constrain the stellar physics of
stars with masses well above $\sim100\,M_\odot$: its isolation makes
it an easier target for observations compared to the similar stars present the crowded core of R136.

The similarities between VFTS682 and the WNh5 stars in the core of
R136 are also in agreement with the ``bully binary'' model of
\cite{fujii:11}. Based on their numerical results, they suggested that
early in the evolution of a cluster, dynamical interactions form an extremely
massive binary, which then tightens its orbit by ejecting other stars passing
by. Interpreting our results for VFTS682 through the lens of their simulations
suggests the presence of a binary with total mass
$M_1+M_2\gtrsim 300\,M_\odot$ in the core of the cluster. Such bully
binary could be R145 according to \cite{fujii:11}.

The kinematic age of VFTS682 puts an
upperlimit to the timescale to form such ``bully binary'' in
R136. The cluster must have been at the very beginning of its
evolution, given the age estimate of $\lesssim 2$\,Myr
\cite{crowther:10,sabbi:12} and the kinematic age of VFTS16.

\todo{check \cite{banerjee} and rewrite next paragraph, they
  have stuff} Unfortunately, their simulations did not include stars more
massive than $100\,M_\odot$, so it is difficult to predict, based on
the existence of VFTS682, how many other stars should have been
ejected from the cluster already and what is their mass
distribution. A comprehensive study of the kinematic properties of the
large sample of stars with extremely large inferred masses visible
around R136 is encouraged.

\cite{lennon:18} have carried out a similar study on VFTS16, another
massive star ($\sim90\,M_\odot$) in the 30 Doradus region, previously
known to be a runaway from the value of
its radial velocity. They also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster. 
The value of $\tau_\mathrm{kin}\simeq0.63$\,Myr we find for VFTS682 (see \Secref{sec:r136_origin}) is smaller
than the corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic age of
$\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

\begin{itemize}
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\end{itemize}

@@

\begin{acknowledgements}
  We are grateful to S.~Torres for extremely useful discussions on the
  Gaia DR2 dataset, and to M.~C.~Ramirez-Tannus for invaluable help.
  The background image of \Figref{fig:main} is based on observations
  made with ESO Telescopes at the La Silla Observatory under programme
  ID 076.C-0888, processed and released by the ESO VOS/ADP group.
  This research made use of APLpy, an open-source plotting package for Python \citep[][]{robitaille:12}.
\end{acknowledgements}
\end{document}
